---
title: "code_robust"
author: "Riccardo Petrella"
date: "2024-02-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Robust statistics 
## Robust Standardization 
```{r}
library(robustbase)
################
#### edit with your data ####
################

oliveoil <- read.table("oliveoil.dat",header=TRUE)
olive <- oliveoil[,3:10]

data <- olive 
################

#### Median and MAD ####
centers <- apply(data, 2, median)
scales <- apply(data, 2, mad)
data_centered <- sweep(data, 2, centers)
data_st <- sweep(data_centered, 2, scales, FUN = "/")
head(data_st)

#### Huber ####
cent_scal <- apply(data, 2, huberM)
centers <- sapply(cent_scal, "[[", 1)
scales <- sapply(cent_scal, "[[", 2)
data_centered <- sweep(data, 2, centers)
data_st <- sweep(data_centered, 2, scales, FUN = "/")
head(data_st)
```
## Parameter estimation 

### One variable 
```{r}
################
#### edit with your data ####
################

data <- dortmund 
variable <- data$birthdeath
huber_const <- 1.5
alpha <- 0.75                                           # For estimation of MCD (Minimum Covariance Determinant)
################
set.seed(12345)

#### Preliminary analytics #### 
## Location 
mean(variable)
sum(variable > mean(variable)) / length(variable)       # Proportion of observations above / below mean 
median(variable) 

## Scale 
sd(variable)
mad(variable, constant = 1.4826)


## Outliers 
which.min(variable)
which.max(variable) 

#### Robust statistics #### 
huberM(variable, k = huber_const)
```

### Whole dataset 

```{r}
################
#### edit with your data ####
################


data <- dortmund 
alpha = 0.5

var_1 <- 2              # Specify variables for visualization
var_2 <- 4              # Specify variables for visualization
################
set.seed(12345)

#### Preliminary analytics #### 
## Location 
colMeans(data)
colMedians(as.matrix(data))

## Scale 
summary(data)

## Correlation 
cor(data[,var_1], data[,var_2])

#### Robust Covariance + Location estimation #### 
## Computation 
mcd_obj <- covMcd(data, alpha = alpha) 

## Location
mcd_obj$center 
mcd_obj$raw.center 

## Covariance 
str(mcd_obj$cov)
str(mcd_obj$raw.cov)

## Correlation
mcd_obj$cov[var_1, var_2] / sqrt(mcd_obj$cov[var_1, var_1] * mcd_obj$cov[var_2, var_2])

#### Visualize ####
plot(data[,c(var_1, var_2)])
## Non-robust
ellipse(colMeans(data[,c(var_1, var_2)]), cov(data)[c(var_1, var_2), c(var_1, var_2)], alpha = 0.01)
## Robust 
ellipse(mcd_obj$center[c(var_1, var_2)], mcd_obj$cov[c(var_1, var_2), c(var_1, var_2)], col = 4, alpha = 0.01)
```

## Outlier Identification 

Outliers by Mahalanobis distance can be 
- Faulty measurements 
- Correct but special 
- Result from non-normality of the process 

Its better to identify outliers from robust statistics which 
are not effected by outliers. 

**Important** 
MCD gives different solutions, depending on the number of variables 
it is computed. This is a main difference to Covariance matrix. 

```{r}
################
#### edit with your data ####
################
data <- read.table("Dortmund_msbd.dat",header=TRUE,row.names=1) 
alpha = 0.75
outlier_level <- 0.99           # When to consider an observation as outlier, based on (normal) Mahanalobis distance 
################
set.seed(12345)

#### Visualize #### 
## Boxplot 
boxplot(scale(data))

## Mahalanobis distance 
mcd_obj <- covMcd(data, alpha = alpha) 
plot(1:nrow(data), sqrt(mcd_obj$mah), ylim = c(0, 30), type = "n", xlab = "Observation", ylab = "Robust Mahalanobis distance")
text(1:nrow(data), sqrt(mcd_obj$mah), rownames(data), cex = 0.7) 
abline(sqrt(qchisq(outlier_level, ncol(data) - 1)), 0, col = 2)
```

## Outliers in regression 

Weights for each observation tell about the importance.
If weight is zero for an observation, it is most certainly an outlier. 
```{r}
################
#### edit with your data ####
################
data <- starsCYG  
formula <- log.light ~ log.Te
# data <- regdata2  
# formula <- y ~ x1 + x2 + x3
var_x <- 2                      # Variable of interest, y should be BEFORE x 
var_y <- 1                      # Position of y in dataset 
################ 
set.seed(12345)

pairs(data) 
#### Non-robust regression ####
## Computation 
lm_norm <- lm(formula, data = data)
summary(lm_norm)

## Diagnostic plots 
par(mfrow = c(2, 2))
plot(lm_norm)
par(mfrow = c(1, 1))

## Visualize regression line 
plot(data[, c(var_x, var_y)])
abline(lm_norm$coef[1], lm_norm$coef[var_x])

#### Robust regression #### 
## Computation
lm_rob <- lmrob(formula, data = data)                   # MM-estimator 
lm_rob <- lmrob(formula, data = data, method = "S")   # S-estimator 

## Diagnostic plots 
summary(lm_rob)
par(mfrow = c(2, 3))
plot(lm_rob)
plot(1:nrow(data), lm_rob$rweights, main = "Observation weights", xlab = "Obs. number", ylab = "Robustness weights")
par(mfrow = c(1, 1))
```